---
title: "02_throw_position_data"
format: html
editor: source
---

#### 0. Importing the dataset

```{r}
rm(list = ls())

library(tidyverse)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")

ufa_throws
```

```{r}
# Flipping the dataset so that positive x is the right side and negative x is the left side
ufa_throws <- ufa_throws |>
  mutate(thrower_x = (-1) * thrower_x,
         receiver_x = (-1) * receiver_x,
         throw_distance = sqrt((receiver_x-thrower_x)^2 + (receiver_y-thrower_y)^2),
         x_diff = receiver_x - thrower_x,
         throw_angle = atan2(y_diff, x_diff))

ufa_throws
```

```{r}
# Filtering for successful throws
successful_throws <- ufa_throws |>
  filter(turnover == 0)

successful_throws
```

#### Main question:

**Which direction of attack (i.e. to the right or left side of the field) has more chances of scoring a goal? Does this vary across teams?**

See the `01_data_exploration.qmd` files for:

- Heat map for the directions of scoring a goal (faceted by quarter)
- Example of a sequence of throws that lead to a goal

Note: A sequence of throws eventually leading to a turnover/goal is (hopefully) uniquely determined by the following variables:

- Possession number
- Game quarter
- Team in possession (`if_home_team`)
- Current scores of home and away teams
- Game ID

#### 1. Number of throws made in a particular wing of the field or crossed the midfield (successful throws only)

```{r}
successful_throws |>
  mutate(throw_path_type = case_when(
    (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
    (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
    (thrower_x >= 0) & (receiver_x <= 0) ~ "right_to_left",
    (thrower_x <= 0) & (receiver_x >= 0) ~ "left_to_right"
  )) |>
  count(throw_path_type) |>
  mutate(prop = n / sum(n))
```

#### 1.1. Additional filter step for successful attacking phases (those that result in a goal)

```{r}
# Create a list of attacking phases that result in a goal
successful_goals_indicators <- ufa_throws |>
  filter(goal == 1) |>
  select(gameID, game_quarter, is_home_team, possession_num,
         home_team_score, away_team_score)

successful_goals_indicators
```

```{r}
# Now only keep these attacking phases in the UFA throws table
successful_attacking_phases <- successful_goals_indicators |>
  left_join(ufa_throws,
            join_by(gameID, game_quarter, is_home_team, possession_num,
                    home_team_score, away_team_score))

successful_attacking_phases
```

#### 2. Number of left-wing, right-wing, and cross-wing passes

Overall distribution across the dataset:

```{r}
successful_attacking_phases <- successful_attacking_phases |>
  mutate(throw_path_type = case_when(
    (thrower_x >= 0) & (receiver_x >= 0) ~ "right_wing",
    (thrower_x <= 0) & (receiver_x <= 0) ~ "left_wing",
    TRUE ~ "cross_throw"
    ),
    team_in_possession = case_when(
      is_home_team == TRUE ~ home_teamID,
      is_home_team == FALSE ~ away_teamID
    ))

successful_attacking_phases
```

```{r}
count_throw_path_type <- successful_attacking_phases |>
  group_by(gameID, game_quarter, team_in_possession, possession_num,
           home_team_score, away_team_score) |>
  # Count the throw path type (which wing?) by each attacking phase in all matches
  count(throw_path_type) |>
  ungroup()

count_throw_path_type
```

```{r}
# Note that we are working on successful attacking phases: the attacks that eventually lead to a goal

throw_path_by_team <- count_throw_path_type |>
  select(team_in_possession, throw_path_type, n) |>
  # Count of different types of throws by team
  group_by(team_in_possession, throw_path_type) |>
  summarize(count_throw_path = n()) |>
  ungroup()

throw_path_by_team

sum_throws_by_team <- count_throw_path_type |>
  select(team_in_possession, n) |>
  # Total of throws by team in possession
  count(team_in_possession)

sum_throws_by_team

prop_throws_by_team <- throw_path_by_team |>
  left_join(sum_throws_by_team, join_by(team_in_possession)) |>
  mutate(prop = count_throw_path / n) |>
  filter(!(team_in_possession %in% c("allstars1", "allstars2")))

prop_throws_by_team
```

```{r}
# Which teams have the highest proportion of cross throws?

prop_throws_by_team |>
  filter(throw_path_type == "cross_throw") |>
  select(team = team_in_possession, prop) |>
  ggplot(aes(y = team, x = prop)) +
  geom_col()
```

#### Other questions (will explore later):

- Which place up the field (y-coordinate) are teams most likely to change their wing of attack?
- How many shots on goal (not passes) are performed inside a wing or cross-wings?
- Number of goals that a team has made after picking up from at least a turnover?
- Explore the starting position (where to build up) and ending position (where there is a shot on goal) of a successful/unsuccessful attacking phase
- Explore the possession time of a successful attacking phase, although this is outside of the throw position data focus of this notebook