---
title: "test-du"
format: html
---

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(forcats)
library(tidytext)
library(ggthemes)

ufa_throws <- read_csv("https://raw.githubusercontent.com/BradenEberhard/Expected-Throwing-Value/refs/heads/main/data/all_games_1024.csv")

ufa_throws <- ufa_throws |> 
  mutate(
    goal = ifelse(receiver_y > 100 & turnover == 0, 1, 0),
    throw_distance = sqrt((receiver_x - thrower_x) ^ 2 + (receiver_y - thrower_y) ^ 2),
    x_diff = receiver_x - thrower_x,
    y_diff = receiver_y - thrower_y,
    throw_angle = atan2(y_diff, x_diff)
  ) |> 
  select(-quarter_point, -total_points, -current_line)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
```

# EDA 

## Game
gameID, possession_num, possession_throw, game_quarter, home_team_score, away_team_score, home_teamID, away_teamID, home_team_win (whether home team wins), score_diff (between home and away teams)

```{r}
ufa_throws |>
  filter(goal == 0 & turnover == 1) # intermediate passing throw, neither turnover nor goal (when turnover = 0, a throw could result in either a goal or a regular passing. When turnover = 1, this means the throw could not result in a goal)

thrower_turnover <- ufa_throws |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_turnover_in_a_season = sum(turnover)) |>
  ungroup() 

thrower_game <- ufa_throws|>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_game_in_a_season = n_distinct(gameID)) |>
  ungroup()

left_join(thrower_turnover, thrower_game, by = c("thrower", "year")) |>
  mutate(avg_turnover_per_game = num_turnover_in_a_season/num_game_in_a_season)
  
thrower_turnover |>
  ggplot(aes(x = avg_turnover_per_game)) +
  geom_histogram() + 
  labs(title = "Distribution of Average Thrower Turnover per Game")
```

```{r}
ufa_throws <- ufa_throws |>
  left_join(thrower_turnover, by = "thrower") |>
  relocate(avg_turnover_per_game)

ufa_throws |>
  group_by(home_teamID) |>
  arrange(desc(avg_turnover_per_game), .by_group = TRUE) |>
  ggplot(aes(x = ))
```

## Individual throwers/receivers
thrower, thrower_x, thrower_y, receiver, receiver_x, receiver_y

## Throw (results in goal/turnover, angle)
turnover, game_quarter, is_home_team, times (game time remaining in seconds), throw_distance, x_diff, y_diff, throw_angle 

## Team metrics
```{r}
ufa_throws <- ufa_throws|>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean))

game_team_status <- ufa_throws |>
  group_by(gameID) |>
  select(home_teamID, away_teamID) |>
  pivot_longer(cols = c("home_teamID", "away_teamID"),
               values_to = "teamID",
               names_to = "status") |>
  slice(1:2) |>
  ungroup() |>
  mutate(status = str_remove(status, "_teamID"))

team_games <- ufa_throws |>
  select(gameID, home_teamID, away_teamID, home_team_win) |>
  group_by(gameID) |>
  slice(1) |>
  pivot_longer(cols = c("home_teamID", "away_teamID"),
               values_to = "teamID",
               names_to = "status") |>
  ungroup() |>
  mutate(status = str_remove(status, "_teamID")) |>
  mutate(win = if_else(status == "home", home_team_win, 1 - home_team_win)) |>
  select(-home_team_win) |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  relocate(gameID, date_clean, year, month, teamID, status, win)
``` 

```{r}
team_games1 <- team_games |>
  group_by(year, teamID) |>
  summarize(n_win = sum(win),
            n_loss = sum(win == 0)) |>
  ungroup()

team_games1 |>
  mutate(teamID = reorder_within(teamID, by = -n_win, within = year, fun = sum, sep = "___")) |>
  pivot_longer(cols = c("n_win", "n_loss"),
               values_to = "n",
               names_to = "result") |>
  mutate(result = str_remove(result, "n_")) |>
  ggplot(aes(x = teamID, y = n, fill = result)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~year, scales = "free_x") + 
  scale_x_reordered() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
all_teams <- unique(ufa_throws$away_teamID) # 26 teams in total 
all_teams <- all_teams[all_teams != "allstars1"]
# cannons, havoc, nitro, outlaws, shred, summit only took part in 2 or 3 seasons 

n_home_games <- ufa_throws |>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  group_by(home_teamID, year) |>
  summarize(n_home_games = n_distinct(gameID))

n_away_games <- ufa_throws |>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  group_by(away_teamID, year) |>
  summarize(n_away_games = n_distinct(gameID)) 

left_join(n_home_games, n_away_games, by = c("home_teamID" = "away_teamID", "year" = "year"))
  
```

```{r}
all_home_games <- ufa_throws |>
  filter(home_teamID %in% all_teams) |>
  group_by(home_teamID, gameID) |>
  summarize(n_throws = n(),
            win = head(home_team_win, n = 1)) |>
  ungroup() 

all_away_games <- ufa_throws |>
  filter(away_teamID %in% all_teams) |>
  group_by(away_teamID, gameID) |>
  summarize(n_throws = n(),
            win = head(home_team_win, n = 1)) |>
  ungroup()

all_home_games |>
  group_by(home_teamID) |>
  count(win) |>
  pivot_wider(names_from = win,
              values_from = n) |>
  rename(loss = "0",
         win = "1")

all_away_games |>
  group_by(away_teamID) |>
  count(win) |>
  pivot_wider(names_from = win,
              values_from = n) |>
  rename(loss = "0",
         win = "1")

all_games |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, home_teamID)
  
# avg turnover rate per game
```


# distance of throws   trajectory of all throws --> goal 


# cluster analysis on player style of play
1. not sure if style of play differs between individual players or combinations of players
## data wrangling 
```{r}
# number of throws/game, number of receives/game, average throw_distance, average throw_y, number of goals per game, (throw_x, throw_angle, x_diff, y_diff)??

# cutter, handler, hybrid 
throw_subset <- ufa_throws |>
  mutate(throwID = row_number()) |>
  relocate(throwID) |>
  select(throwID, thrower, receiver, throw_distance, thrower_y, gameID, goal)

player_n_throw <- throw_subset |>
  group_by(thrower, gameID) |>
  count() |>
  ungroup() |>
  group_by(thrower) |>
  summarize(n_throw = sum(n),
            n_game_throw = n_distinct(gameID)) |>
  mutate(avg_n_throw = n_throw/n_game_throw) |>
  rename(player = thrower)

player_n_receive <- throw_subset |>
  group_by(receiver, gameID) |>
  count() |>
  ungroup() |>
  group_by(receiver) |>
  summarize(n_receive = sum(n),
            n_game_receive = n_distinct(gameID)) |>
  mutate(avg_n_receive = n_receive/n_game_receive) |>
  rename(player = receiver)

player_n <- left_join(player_n_throw, player_n_receive, by = "player") |>
  select(player, avg_n_throw, avg_n_receive) 
```

still need to get these metrics for each player: average throw_distance, average throw_y, number of goals per game

```{r}
player_throw_dist <- throw_subset |>
  group_by(thrower) |>
  summarize(avg_throw_distance = sum(throw_distance)/n(),
            avg_thrower_y = sum(thrower_y)/n()) |>
  ungroup() |>
  rename(player = thrower)

player_n <- left_join(player_n, player_throw_dist, by = "player")
```

```{r}
player_n_goal <- throw_subset |>
  group_by(gameID, thrower) |>
  summarize(n_goal = sum(goal)) |>
  ungroup() |>
  group_by(thrower) |>
  summarize(avg_n_goal = sum(n_goal)/n()) |>
  ungroup() |>
  rename(player = thrower)

player_n <- left_join(player_n, player_n_goal, by = "player")
```

## 1. K-means clustering
```{r}
std_player_stats <- player_n |>
  select(-player) |>
  scale(center = TRUE, scale = TRUE) |>
  as_tibble() |>
  mutate(avg_n_receive = coalesce(avg_n_receive, 0))

std_player_stats1 <- player_n |>
  mutate(across(where(is.numeric), ~ as.numeric(scale(.)))) |>
  mutate(avg_n_receive = coalesce(avg_n_receive, 0))
```

```{r}
# elbow plot to determine the number of clusters 
player_kmeans <- function(k) {
  kmeans_results <- std_player_stats |>
    #select(avg_n_throw, avg_throw_distance) |>
    kmeans(centers = k, nstart = 30, iter.max = 30)
  
  kmeans_out <- tibble(
    clusters = k,
    total_wss = kmeans_results$tot.withinss
  )
  return(kmeans_out)
}

n_cluster_search <- 1:6

kmeans_search <- n_cluster_search |>
  map(player_kmeans) |>
  bind_rows()

kmeans_search |> 
  ggplot(aes(x = clusters, y = total_wss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = n_cluster_search)
```

```{r}
set.seed(743932)
player_kmeans1 <- std_player_stats |>
  #select(avg_n_throw, avg_throw_distance) |>
  kmeans(algorithm = "Hartigan-Wong",
         centers = 3,
         nstart = 30,
         iter.max = 30)

library(ggimage)
std_player_stats |>
  mutate(player_clusters = as.factor(player_kmeans1$cluster)) |>
  ggplot(aes(x = avg_n_throw, y = avg_throw_distance, color = player_clusters)) +
  geom_point() +
  scale_color_colorblind(labels = c("1" = "Hybrid",
                                    "2" = "Cutter",
                                    "3" = "Handler")) +
  theme_minimal() +
  coord_fixed() +
  annotation_raster(raster = img,
                    xmin = 2.2, xmax = 2.6,
                    ymin = 2.6, ymax = 3.0)
  # geom_image(
  #   data = img_data,
  #   aes(x = x_coord, y = y_coord, image = image_path),
  #   size = 0.8
  # )
```

```{r}
# "gsanner", "aroy": cutters
# "pjanas", "bmatzuka": handlers (or just greater thrower?)
# "jwilliams", "etaylor", "acoffin", "jkerr", "rbarker", "khenke", "tmonroe": hybrids 

std_player_stats1 |>
  filter(player %in% c("jwilliams", "etaylor", "acoffin", "jkerr", "rbarker", "khenke", "tmonroe", "gsanner", "aroy", "pjanas", "bmatzuka"))

img_data <- tibble(
  x_coord = 2.4130366,
  y_coord = 2.82300490,
  image_path = "~/Desktop/Carleton Docs/R/CMSAC2026/eda project/ufa-throwing-cmsacamp-2026/Du/jwilliams_profile.jpg")

img <- readJPEG("~/Desktop/Carleton Docs/R/CMSAC2026/eda project/ufa-throwing-cmsacamp-2026/Du/jwilliams_profile.jpg")
```


## 1. Hierarchical clustering
```{r}
players_dist <- std_player_stats |>
  # select(avg_n_throw, avg_throw_distance) |>
  dist()

players_dist_matrix <- as.matrix(players_dist)
rownames(players_dist_matrix) <- std_player_stats1$player
colnames(players_dist_matrix) <- std_player_stats1$player

players_dist_matrix[1:4, 1:4]
```
```{r}
library(factoextra)
std_player_stats |>
  fviz_nbclust(FUN = hcut, method = "silhouette")
```

```{r}
players_complete <- players_dist |>
  hclust(method = "complete")

std_player_stats1 |>
  mutate(cluster = as.factor(cutree(players_complete, k = 3))) |>
  relocate(cluster) |>
  ggplot(aes(x = avg_n_throw, y = avg_throw_distance, color = cluster)) +
  geom_point() +
  coord_fixed() +
  scale_color_colorblind(labels = c("1" = "Handler",
                                    "2" = "Hybrid",
                                    "3" = "Cutter")) +
  theme_minimal()
```

```{r}
library(ggdendro)

players_complete |>
  ggdendrogram(labels = FALSE, leaf_labels = FALSE) +
  labs(y = "Dissimilarity between clusters") + 
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank())
```

## 3. Soft clustering with GMM 

```{r}
library(mclust)

player_mclust <- std_player_stats |>
  select(avg_n_throw, avg_throw_distance) |>
  Mclust()

summary(player_mclust)
```

```{r}
library(gt)
library(broom)
gt(tidy(player_mclust))
```


2. Cluster throws on time remaining, score difference 
compare completion rate, throw distance

adjust shape/alpha based on whether there was a turnover?

## 2. K-means clustering

```{r}
ufa_throws$turnover <- as.factor(ufa_throws$turnover)
std_throw_stats <- ufa_throws |>
  filter(year == 2024 & month == 8) |>
  mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
```

```{r}
# elbow plot 
std_throw_stats |>
  select(times, score_diff) |>
  fviz_nbclust(kmeans, method = "wss")
```

```{r}
set.seed(79174)
throw_kmeans <- std_throw_stats |>
  select(score_diff, throw_distance) |>
  kmeans(algorithm = "Hartigan-Wong",
         centers = 3, 
         nstart = 30, 
         iter.max = 30)

std_throw_stats <- std_throw_stats |>
  mutate(throw_clusters = as.factor(throw_kmeans$cluster))

std_throw_stats |>
  ggplot(aes(x = times, y = score_diff, color = throw_clusters)) +
  geom_point() +
  scale_color_colorblind() +
  coord_fixed() +
  theme_minimal()

std_throw_stats |>
  mutate(throw_clusters = as.factor(throw_kmeans$cluster)) |>
  ggplot(aes(x = score_diff, y = throw_distance, color = throw_clusters)) +
  geom_point() + 
  geom_hline(yintercept = 1.60, color = "red", linetype = "dashed") +
  scale_color_colorblind() +
  coord_fixed() +
  theme_minimal()

# how can we interpret these plots, especially with "score_diff" being discrete and "turnover" being a indicator variable?

# what's considered a long-distance throw: ~1.6 sd from the mean and above (~35 yd)
1.6*sd(ufa_throws$throw_distance) + mean(ufa_throws$throw_distance)
```

```{r}
std_throw_stats |>
  ggplot(aes(x = throw_distance, color = turnover)) +
  geom_dotplot() +
  facet_wrap(~throw_clusters) +
  theme_minimal()
```


## 2. Hierarchical clustering
```{r}
std_throw_stats <- ufa_throws |>
  filter(year == 2024) |>
  mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
```

```{r}
throws_dist <- std_throw_stats |>
  select(times, score_diff) |>
  dist()

# Error: vector memory limit of 16.0 Gb reached, see mem.maxVSize()
library(amap)
throws_dist <- Dist(std_throw_stats, method = "euclidean")
```


# Home vs. away differences 
```{r}
team_avg_net <- full_team_stats |>
  select(team, home_games, away_games) |>
  left_join(home_score, by = "team") |>
  rename(total_goals_scored_home = total_goals_scored,
         total_goals_conceded_home = total_goals_conceded) |>
  left_join(away_score, by = "team") |>
  rename(total_goals_scored_away = total_goals_scored,
         total_goals_conceded_away = total_goals_conceded) |>
  mutate(avg_goal_scored_home = total_goals_scored_home/home_games,
         avg_goal_conceded_home = total_goals_conceded_home/home_games,
         avg_goal_scored_away = total_goals_scored_away/away_games,
         avg_goal_conceded_away = total_goals_conceded_away/away_games,
         avg_net_home = avg_goal_scored_home - avg_goal_conceded_home,
         avg_net_away = avg_goal_scored_away - avg_goal_conceded_away) |>
  select(team, avg_net_home, avg_net_away) |>
  pivot_longer(cols = c("avg_net_home", "avg_net_away"),
               values_to = "avg_net",
               names_to = "status") |>
  mutate(status = str_remove(status, "avg_net_"))
```

```{r}
# spaghetti_plot

ggplot(team_avg_net, aes(x = status, y = avg_net)) +
  geom_point() +
  geom_line(aes(group = team), color = "darkgray", alpha = 0.4) +
  theme_minimal()
```

