---
title: "test-du"
format: html
---

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(forcats)
library(tidytext)

ufa_throws <- read_csv("https://raw.githubusercontent.com/BradenEberhard/Expected-Throwing-Value/refs/heads/main/data/all_games_1024.csv")

ufa_throws <- ufa_throws |> 
  mutate(
    goal = ifelse(receiver_y > 100 & turnover == 0, 1, 0),
    throw_distance = sqrt((receiver_x - thrower_x) ^ 2 + (receiver_y - thrower_y) ^ 2),
    x_diff = receiver_x - thrower_x,
    y_diff = receiver_y - thrower_y,
    throw_angle = atan2(y_diff, x_diff)
  ) |> 
  select(-quarter_point, -total_points, -current_line)

ufa_throws <- read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ufa_throws.csv")
```

# EDA 

## Game
gameID, possession_num, possession_throw, game_quarter, home_team_score, away_team_score, home_teamID, away_teamID, home_team_win (whether home team wins), score_diff (between home and away teams)

```{r}
ufa_throws |>
  filter(goal == 0 & turnover == 1) # intermediate passing throw, neither turnover nor goal (when turnover = 0, a throw could result in either a goal or a regular passing. When turnover = 1, this means the throw could not result in a goal)

thrower_turnover <- ufa_throws |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_turnover_in_a_season = sum(turnover)) |>
  ungroup() 

thrower_game <- ufa_throws|>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, thrower) |>
  summarize(num_game_in_a_season = n_distinct(gameID)) |>
  ungroup()

left_join(thrower_turnover, thrower_game, by = c("thrower", "year")) |>
  mutate(avg_turnover_per_game = num_turnover_in_a_season/num_game_in_a_season)
  
thrower_turnover |>
  ggplot(aes(x = avg_turnover_per_game)) +
  geom_histogram() + 
  labs(title = "Distribution of Average Thrower Turnover per Game")
```

```{r}
ufa_throws <- ufa_throws |>
  left_join(thrower_turnover, by = "thrower") |>
  relocate(avg_turnover_per_game)

ufa_throws |>
  group_by(home_teamID) |>
  arrange(desc(avg_turnover_per_game), .by_group = TRUE) |>
  ggplot(aes(x = ))
```

## Individual throwers/receivers
thrower, thrower_x, thrower_y, receiver, receiver_x, receiver_y

## Throw (results in goal/turnover, angle)
turnover, game_quarter, is_home_team, times (game time remaining in seconds), throw_distance, x_diff, y_diff, throw_angle 

## Team metrics
```{r}
ufa_throws <- ufa_throws|>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean))

game_team_status <- ufa_throws |>
  group_by(gameID) |>
  select(home_teamID, away_teamID) |>
  pivot_longer(cols = c("home_teamID", "away_teamID"),
               values_to = "teamID",
               names_to = "status") |>
  slice(1:2) |>
  ungroup() |>
  mutate(status = str_remove(status, "_teamID"))

team_games <- ufa_throws |>
  select(gameID, home_teamID, away_teamID, home_team_win) |>
  group_by(gameID) |>
  slice(1) |>
  pivot_longer(cols = c("home_teamID", "away_teamID"),
               values_to = "teamID",
               names_to = "status") |>
  ungroup() |>
  mutate(status = str_remove(status, "_teamID")) |>
  mutate(win = if_else(status == "home", home_team_win, 1 - home_team_win)) |>
  select(-home_team_win) |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  relocate(gameID, date_clean, year, month, teamID, status, win)
``` 

```{r}
team_games1 <- team_games |>
  group_by(year, teamID) |>
  summarize(n_win = sum(win),
            n_loss = sum(win == 0)) |>
  ungroup()

team_games1 |>
  mutate(teamID = reorder_within(teamID, by = -n_win, within = year, fun = sum, sep = "___")) |>
  pivot_longer(cols = c("n_win", "n_loss"),
               values_to = "n",
               names_to = "result") |>
  mutate(result = str_remove(result, "n_")) |>
  ggplot(aes(x = teamID, y = n, fill = result)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~year, scales = "free_x") + 
  scale_x_reordered() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
all_teams <- unique(ufa_throws$away_teamID) # 26 teams in total 
all_teams <- all_teams[all_teams != "allstars1"]
# cannons, havoc, nitro, outlaws, shred, summit only took part in 2 or 3 seasons 

n_home_games <- ufa_throws |>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  group_by(home_teamID, year) |>
  summarize(n_home_games = n_distinct(gameID))

n_away_games <- ufa_throws |>
  filter(!(home_teamID %in% c("allstars1", "allstars2") & away_teamID %in% c("allstars1", "allstars2"))) |>
  group_by(away_teamID, year) |>
  summarize(n_away_games = n_distinct(gameID)) 

left_join(n_home_games, n_away_games, by = c("home_teamID" = "away_teamID", "year" = "year"))
  
```

```{r}
all_home_games <- ufa_throws |>
  filter(home_teamID %in% all_teams) |>
  group_by(home_teamID, gameID) |>
  summarize(n_throws = n(),
            win = head(home_team_win, n = 1)) |>
  ungroup() 

all_away_games <- ufa_throws |>
  filter(away_teamID %in% all_teams) |>
  group_by(away_teamID, gameID) |>
  summarize(n_throws = n(),
            win = head(home_team_win, n = 1)) |>
  ungroup()

all_home_games |>
  group_by(home_teamID) |>
  count(win) |>
  pivot_wider(names_from = win,
              values_from = n) |>
  rename(loss = "0",
         win = "1")

all_away_games |>
  group_by(away_teamID) |>
  count(win) |>
  pivot_wider(names_from = win,
              values_from = n) |>
  rename(loss = "0",
         win = "1")

all_games |>
  mutate(date_clean = as_date(str_sub(gameID, 1, 10)),
         year = year(date_clean),
         month = month(date_clean)) |>
  group_by(year, home_teamID)
  
# avg turnover rate per game
```


# distance of throws   trajectory of all throws --> goal 